---
title: "Appendix"
author: "Max Aantjes"
date: "15/09/2020"
output:
  github_document:
    toc: true
    toc_depth: 2
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The following sections contain all computations for the creation of the exploratory data results summarised in the brief paper. Computations are accompanied by references and justifications to clarify any choices made. 

# R Packages
The following R packages are used in the subsequent code sections. 
```{r, message = FALSE}
library(openxlsx)
library(stringr)
library(tidyverse)
library(xts)
library(foreign)
```
The following code loads a useful function to account for special characters.
```{r}
rep_sp_char <- function(x){
        sp_char <- c("á", "é", "í", "ó", "ñ")
        rep <- c("a", "e", "i", "o", "n")
        for(i in 1:length(rep)){
                x <- gsub(sp_char[i], rep[i], x)}
        return(x)}
```

# Data Extraction
## Start-of-school-year Matriculations (2009-2019)
### Data Source and Relevance
The data used is summarised matriculation data calculated from the AMIE survey conducted by INEC. This is an annual survey which collects census data on the number of students and teachers at each Ecuadorian primary and secondary institution. Code books and methodology documents are available in this [online repository](https://educacion.gob.ec/amie/)

These data sets are updated every year, so it possible that links stop working. The relevant data sets (tabulados por sostenimiento) can be downloaded from this website: https://educacion.gob.ec/indice-de-tabulados/.

### Download
The following code downloads the data. 
```{r echo = TRUE, results = 'hide', message = FALSE}
## Download Data
links <- list("https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15555", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15556", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15557", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15558")
dfl <- lapply(links, read.xlsx, sheet = 5, startRow = 12)
addcol <- function(x){
        y <- c("fiscal", "particular", "municipal", "fiscomisional")
        for(i in 1:length(x)){x[[i]]$type <- y[i]}
        return(x)}
dfl1 <- addcol(dfl)
dfl2 <- lapply(dfl1, pivot_longer, cols = "2009-2010.Inicio":"2019-2020.Inicio",
               names_to = "schoolyear", values_to = "no.students")
df <- do.call(rbind, dfl2) %>%
        rename("province" = Provincia)
```

## Start-of-school-year Matriculations (2020)

```{r}
sdf <- read.csv("dfzone3.csv") %>%
        mutate(schoolyear = "2020-2021.Inicio") %>%
        pivot_longer(cols = "fiscal":"municipal", names_to = "type",
                     values_to = "no.students")
```

## Population projections by year
https://www.ecuadorencifras.gob.ec/proyecciones-poblacionales/
```{r}
url <- "https://www.ecuadorencifras.gob.ec/documentos/web-inec/Poblacion_y_Demografia/Proyecciones_Poblacionales/PROYECCION_POR_EDADES_PROVINCIAS_2010-2020_Y_NACIONAL_2010-2020.xlsx"
sheets <- 2:12
years <- 2010:2020
read.mult.xlsx <- function(x, y, z){
  q <- list()
  for(i in 1:length(y)){
        q[[i]] <- read.xlsx(x, sheet = y[i], startRow = 8)
        q[[i]]$year = z[i]}
  q <- do.call(rbind, q)
  return(q)}
tdf <- read.mult.xlsx(url, sheets, years)
```

## 

# Data Cleaning
## Year on Year data of start-of-school-year Matriculations
```{r}
## Merge data 
df1 <- rbind(df, sdf)
```

```{r}
## Create a time series data frame
df2 <- df1 %>%
        mutate(schoolyear = as.Date(paste0(
                str_extract(schoolyear, "^[0-9]{4}"), 
                "-09-01"), format = "%Y-%m-%d"))
        
```

```{r}
## Select data of interest
#pvint <- c("chimborazo", "cotopaxi", "pastaza", "tungurahua")
dftype <- df2 %>%
        mutate(province = tolower(rep_sp_char(province)))# %>%
        #filter(province %in% pvint) 

## Make data frame with total amount of students
dftot <- dftype %>%
          group_by(province, schoolyear) %>%
          summarise(matriculated.students = sum(no.students)) %>%
          ungroup()
```



## Population projections by year
```{r}
## Clean column names
names(tdf) <- tolower(rep_sp_char(names(tdf)))

## Select data of interest
tdf1 <- tdf %>%
        mutate(age = gsub(" ", "", x1)) %>%
        filter(age %in% c("5-9", "10-14", "15-19")) %>%
        mutate(schoolyear = as.Date(
                paste0(year, "-09-01"), format = "%Y-%m-%d")) %>%
        select(schoolyear, age, "chimborazo" = 'chimbo-razo', cotopaxi, 
               pastaza, "total nacional" = "total.pais",
               "tungurahua" = 'tungu-rahua', 
               ) %>%
        group_by(schoolyear) %>%
        summarise_at(vars(chimborazo:tungurahua), sum) %>%
        pivot_longer(cols = chimborazo:tungurahua, 
                     values_to = "estimated.5-19.yo", 
                     names_to = "province") %>%
        ungroup()
```

```{r}
dftot1 <- dftot%>%
        merge(tdf1) %>%
        pivot_longer(cols = 3:4, values_to = "population.size",
                     "population.type")
```

## Exploratory Data analysis
# Country-Level
```{r}
dftot2 <- dftot1 %>%
            filter(province == "total nacional")
p <- ggplot(dat = dftot2, aes(y = population.size, x = schoolyear,
                              col = population.type))
p1 <- p + labs(x = "school year", "population size",
               title = "Number of students enrolled at start of school year\nvs. estimated population aged 5-19 y/o")
p2 <- p1 + geom_point() + geom_line()
p3 <- p2 + theme_bw()
p3
```

```{r}
dftot2 <- dftot1 %>%
            filter(province != "total nacional") %>%
            mutate(population.type = ifelse(
            population.type == "matriculated.students", 
            "Matriculated Students", "5-19 Year-Olds (Estimate)"))
p <- ggplot(dat = dftot2, aes(y = population.size/1000, x = schoolyear,
                              col = population.type))
p1 <- p + facet_grid(~province)
p2 <- p1 + geom_point() + geom_line() + geom_vline(aes(xintercept = as.Date("13-03-2020", format = "%d-%m-%Y"), col = "Start School Lockdown"))
p3 <- p2 + theme_bw()
p4 <- p3 + labs(title = "Number of secondary and primary students matriculated\nat the start of school year vs. estimated population\naged 5-19 years old",
                x = "Year", y = "Population Size (in thousands)") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p5 <- p4 + scale_colour_discrete(name = "")
p5
```

```{r}
df1 <- df %>%
        select("sum.students" = 'total.estudiantes.(inicial.a.3ro..de.bachillerato)', 
               "sum.teachers" ="total.docentes", 
               "school.name" = nombre.institucion, 
               "area" = zona.inec, "finance.type" = sostenimiento, canton, 
               "province" = provincia, "zone" = zona, 
               "education.level" = nivel.educacion,
               "region" = regimen.escolar, modallidad) %>%
        mutate(finance.type = factor(tolower(finance.type))) %>%
        mutate(area = factor(str_remove(area, "INEC"))) %>%
        mutate(canton = factor(tolower(rep_sp_char(canton)))) %>%
        mutate(province = factor(tolower(rep_sp_char(province))))

sdf1 <- sdf %>%
        pivot_longer(cols = c(fiscal, fiscomisional, particular, municipal),
                     values_to = "sum.students", names_to = "finance.type") 
```

```{r}
df2 <- df1 %>%
        group_by(finance.type) %>%
        summarise(sum(sum.students))
sum(df1$sum.students)
```


```{r}
df2 <- df1 %>%
        filter(zone == "ZONA 3") %>%
        filter(finance.type %in% c("fiscal", "particular"))

sdf2 <- sdf1 %>%
        filter(finance.type %in% c("fiscal", "particular"))

## Calculate Statistics
df3 <- df2 %>%
        group_by(province, finance.type) %>%
        summarise(sum.students = sum(sum.students))

## Join the two tables
df4 <- df3 %>%
        full_join(sdf2, by = c("province", "finance.type")) %>%
        rename(y1 = 3, y2 = 4) %>%
        pivot_longer(cols = c(y1, y2), values_to = "sum.students",
                     names_to = "schoolyear")

## Calculate change
df5 <- df4 %>%
        pivot_wider(values_from = sum.students, names_from = schoolyear) %>%
        mutate(change = (y2-y1))

## Plot
p <- ggplot(data = df5, aes(y = change, 
                            x = province, fill = finance.type))
p1 <- p + geom_bar(stat = "identity", position = "dodge")
p2 <- p1 + theme_bw() +
        theme(axis.text.x=element_text(angle = 90, vjust = 0.5))
p3 <- p2 + labs(y = "Change in the number of Students (Absolute)",
                x = "Province",
                title = "Figure 1: The Change in Primary and Secondary\nEducation Enrolment between 2019 and 2020 in Zone 3",
                caption = "Source: Data from INEC (AMIE) and MINEDUC\nData Summary by author (Max Aantjes)")
p4 <- p3 + scale_fill_discrete(name = "School Type", labels = c("Public",
                                          "Private"))
p4


```

```{r}
link <- "https://www.ecuadorencifras.gob.ec//documentos/datos/Poblacion_y_Demografia/nacimientos_defunciones/nacimientos/spss/dbd_nacidos_vivos_spss_2015.zip"

```

