---
title: "Appendix"
author: "Max Aantjes"
date: "15/09/2020"
output:
  github_document:
    toc: true
    toc_depth: 2
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The following sections contain all computations for the creation of the exploratory data results summarised in the brief paper. Computations are accompanied by references and justifications to clarify any choices made. 

# R Packages
The following R packages are used in the subsequent code sections. 
```{r, message = FALSE}
library(openxlsx)
library(stringr)
library(tidyverse)
library(xts)
library(foreign)
```
The following code loads a useful function to account for special characters.
```{r}
rep_sp_char <- function(x){
        sp_char <- c("á", "é", "í", "ó", "ñ")
        rep <- c("a", "e", "i", "o", "n")
        for(i in 1:length(rep)){
                x <- gsub(sp_char[i], rep[i], x)}
        return(x)}
```

# Data Extraction
## Start-of-school-year Matriculations (2009-2019)
### Data Source and Relevance
The data used is summarised matriculation data calculated from the AMIE survey conducted by INEC. This is an annual survey which collects census data on the number of students and teachers at each Ecuadorian primary and secondary institution. Code books and methodology documents are available in this [online repository](https://educacion.gob.ec/amie/)

These data sets are updated every year, so it possible that links stop working. The relevant data sets (tabulados por sostenimiento) can be downloaded from this website: https://educacion.gob.ec/indice-de-tabulados/.

### Download
The following code downloads the data. 
```{r echo = TRUE, results = 'hide', message = FALSE}
## Download Data
links <- list("https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15555", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15556", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15557", "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15558")
dfl <- lapply(links, read.xlsx, sheet = 5, startRow = 12)
addcol <- function(x){
        y <- c("fiscal", "particular", "municipal", "fiscomisional")
        for(i in 1:length(x)){x[[i]]$type <- y[i]}
        return(x)}
dfl1 <- addcol(dfl)
dfl2 <- lapply(dfl1, pivot_longer, cols = "2009-2010.Inicio":"2019-2020.Inicio",
               names_to = "schoolyear", values_to = "no.students")
df <- do.call(rbind, dfl2) %>%
        rename("province" = Provincia)
```

## Start-of-school-year Matriculations Zone 3 (2020)
### Data Source and Relevance
This data summarises the matriculation data in four Ecuadorian provinces at the start of the 2020-2021 school year. The data was taken from a news update by the Ecuadorian Ministry of education. I assume this data to be as accurate as the AMIE tabulated data; this is unlikely to hold true, but it's the best public data available. As the small table was presented as an image, I tabulated it myself in excel and uploaded it into the GitHub repository. Both forms of the data are accessed as follows:  

* Download the excel file [here](https://github.com/MaxAantjes/Exp-Analysis-Dropout-EDU-EC-COVID19/blob/master/dfzone3.csv)  
* Read the news article [here](https://educacion.gob.ec/374-908-estudiantes-inician-un-nuevo-ano-lectivo-en-la-zona-3/)  

### Download
The following code downloads the data. 
```{r}
sdf <- read.csv("dfzone3.csv") %>%
        mutate(schoolyear = "2020-2021.Inicio") %>%
        pivot_longer(cols = "fiscal":"municipal", names_to = "type",
                     values_to = "no.students")
```

## Population projections by year
### Data Source and Relevance
The population projects were calculated by INEC based on the 2010 population census. The methodology and code book is available in the relevant (government repository)[https://www.ecuadorencifras.gob.ec/proyecciones-poblacionales/].

### Download
The following code downloads the data.
```{r}
url <- "https://www.ecuadorencifras.gob.ec/documentos/web-inec/Poblacion_y_Demografia/Proyecciones_Poblacionales/PROYECCION_POR_EDADES_PROVINCIAS_2010-2020_Y_NACIONAL_2010-2020.xlsx"
sheets <- 2:12
years <- 2010:2020
read.mult.xlsx <- function(x, y, z){
  q <- list()
  for(i in 1:length(y)){
        q[[i]] <- read.xlsx(x, sheet = y[i], startRow = 8)
        q[[i]]$year = z[i]}
  q <- do.call(rbind, q)
  return(q)}
tdf <- read.mult.xlsx(url, sheets, years)
```

## AMIE Data (2019)
### Data Source and Relevance
The data used is the AMIE survey conducted by INEC (non-summarised). This annual survey collects census data on the number of students and teachers at each Ecuadorian primary and secondary institution. Code books and methodology documents are available in this [online repository](https://educacion.gob.ec/amie/)

### Download
The following code downloads the data.
```{r}
link <- "https://educacion.gob.ec/wp-content/plugins/download-monitor/download.php?id=15713"
fdf <- read.xlsx(link,  sheet = "Registros Administrativos", startRow = 13)
names(fdf)[2] <- "duplicate" #Rename duplicated column to ensure tidyverse works
names(fdf) <- tolower(rep_sp_char(names(fdf)))
```

# Data Cleaning
## Year on Year data of start-of-school-year Matriculations
The following code merges the data from the AMIE summarised data (2009-2019) and the government announcement (2020). It then transforms the data set into a time series data frame, calculates the total amount of students per year and selects the four provinces of interest (Chimborazo, Cotopaxi, Pastaza and Tungurahua). It also clarifies the names of the type of education.  
```{r}
pvint <- c("chimborazo", "cotopaxi", "pastaza", "tungurahua")
df1 <- rbind(df, sdf) %>%
        mutate(schoolyear = as.Date(paste0(
                str_extract(schoolyear, "^[0-9]{4}"), 
                "-09-01"), format = "%Y-%m-%d"))  %>%
        mutate(province = factor(tolower(rep_sp_char(province)))) %>%
        filter(province %in% pvint) %>%
        mutate(type = factor(type, levels = c("fiscal", "particular",
                                              "municipal", "fiscomisional"),
                             labels = c("public", "private", "municipal",
                                        "mixed")))
head(df1)
```
This results in the following codebook: 
```{r, echo = FALSE}
colnames <- c("province", "type", "schoolyear", "no.students")
explanation <- c("Ecuadorian province name.", "Indicates what finance type the group of education institutions belongs to.", "Approximate Date of School Year Start", "Number of students enrolled at the start of the school year")
values <- c("factor of 4 levels: chimborazo, ...", "factor of 4 levels: public, ...", "Date", "Numeric")
table <- data.frame(cbind(colnames, explanation, values))
kable(table, caption = "Code Book Table 1")

```

## Population projections by year
```{r}
## Clean column names
names(tdf) <- tolower(rep_sp_char(names(tdf)))

## Select data of interest
tdf1 <- tdf %>%
        mutate(age = gsub(" ", "", x1)) %>%
        filter(age %in% c("5-9", "10-14", "15-19")) %>%
        mutate(schoolyear = as.Date(
                paste0(year, "-09-01"), format = "%Y-%m-%d")) %>%
        select(schoolyear, age, "chimborazo" = 'chimbo-razo', cotopaxi, 
               pastaza, "total nacional" = "total.pais",
               "tungurahua" = 'tungu-rahua', 
               ) %>%
        group_by(schoolyear) %>%
        summarise_at(vars(chimborazo:tungurahua), sum) %>%
        pivot_longer(cols = chimborazo:tungurahua, 
                     values_to = "estimated.5-19.yo", 
                     names_to = "province") %>%
        ungroup()
```

```{r}
dftot1 <- dftot%>%
        merge(tdf1) %>%
        pivot_longer(cols = 3:4, values_to = "population.size",
                     "population.type")
```

## Exploratory Data analysis
# Country-Level
```{r}
dftot2 <- dftot1 %>%
            filter(province == "total nacional")
p <- ggplot(dat = dftot2, aes(y = population.size, x = schoolyear,
                              col = population.type))
p1 <- p + labs(x = "school year", "population size",
               title = "Number of students enrolled at start of school year\nvs. estimated population aged 5-19 y/o")
p2 <- p1 + geom_point() + geom_line()
p3 <- p2 + theme_bw()
p3
```

```{r}
dftot2 <- dftot1 %>%
            filter(province != "total nacional") %>%
            mutate(population.type = ifelse(
            population.type == "matriculated.students", 
            "Matriculated Students", "5-19 Year-Olds (Estimate)"))
p <- ggplot(dat = dftot2, aes(y = population.size/1000, x = schoolyear,
                              col = population.type))
p1 <- p + facet_grid(~province)
p2 <- p1 + geom_point() + geom_line() + geom_vline(aes(xintercept = as.Date("13-03-2020", format = "%d-%m-%Y"), col = "Start School Lockdown"))
p3 <- p2 + theme_bw()
p4 <- p3 + labs(title = "Number of secondary and primary students matriculated\nat the start of school year vs. estimated population\naged 5-19 years old",
                x = "Year", y = "Population Size (in thousands)") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p5 <- p4 + scale_colour_discrete(name = "")
p5
```

```{r}
df1 <- df %>%
        select("sum.students" = 'total.estudiantes.(inicial.a.3ro..de.bachillerato)', 
               "sum.teachers" ="total.docentes", 
               "school.name" = nombre.institucion, 
               "area" = zona.inec, "finance.type" = sostenimiento, canton, 
               "province" = provincia, "zone" = zona, 
               "education.level" = nivel.educacion,
               "region" = regimen.escolar, modallidad) %>%
        mutate(finance.type = factor(tolower(finance.type))) %>%
        mutate(area = factor(str_remove(area, "INEC"))) %>%
        mutate(canton = factor(tolower(rep_sp_char(canton)))) %>%
        mutate(province = factor(tolower(rep_sp_char(province))))

sdf1 <- sdf %>%
        pivot_longer(cols = c(fiscal, fiscomisional, particular, municipal),
                     values_to = "sum.students", names_to = "finance.type") 
```

```{r}
df2 <- df1 %>%
        group_by(finance.type) %>%
        summarise(sum(sum.students))
sum(df1$sum.students)
```


```{r}
df2 <- df1 %>%
        filter(zone == "ZONA 3") %>%
        filter(finance.type %in% c("fiscal", "particular"))

sdf2 <- sdf1 %>%
        filter(finance.type %in% c("fiscal", "particular"))

## Calculate Statistics
df3 <- df2 %>%
        group_by(province, finance.type) %>%
        summarise(sum.students = sum(sum.students))

## Join the two tables
df4 <- df3 %>%
        full_join(sdf2, by = c("province", "finance.type")) %>%
        rename(y1 = 3, y2 = 4) %>%
        pivot_longer(cols = c(y1, y2), values_to = "sum.students",
                     names_to = "schoolyear")

## Calculate change
df5 <- df4 %>%
        pivot_wider(values_from = sum.students, names_from = schoolyear) %>%
        mutate(change = (y2-y1))

## Plot
p <- ggplot(data = df5, aes(y = change, 
                            x = province, fill = finance.type))
p1 <- p + geom_bar(stat = "identity", position = "dodge")
p2 <- p1 + theme_bw() +
        theme(axis.text.x=element_text(angle = 90, vjust = 0.5))
p3 <- p2 + labs(y = "Change in the number of Students (Absolute)",
                x = "Province",
                title = "Figure 1: The Change in Primary and Secondary\nEducation Enrolment between 2019 and 2020 in Zone 3",
                caption = "Source: Data from INEC (AMIE) and MINEDUC\nData Summary by author (Max Aantjes)")
p4 <- p3 + scale_fill_discrete(name = "School Type", labels = c("Public",
                                          "Private"))
p4


```

```{r}
link <- "https://www.ecuadorencifras.gob.ec//documentos/datos/Poblacion_y_Demografia/nacimientos_defunciones/nacimientos/spss/dbd_nacidos_vivos_spss_2015.zip"

```

